Deep Learning for 3D Brain Activity Scans

### 2019-12-03 paper/submission.tex
Working through paper, up to results looks pretty good

\author{Evan Albright\\
University of Massachusetts Amherst\\
140 Governors Dr, Amherst, MA 01002\\
{\tt\small ealbright@umass.edu}
}

@@

%-------------------------------------------------------------------------
\begin{abstract}
   Abstract placeholder
\end{abstract}

@@

Classification of activity and structures should come quite naturally with deep learning practices.
Previous works include the development of Restricted Boltzmann Machines, and Deep Belief Networks~\cite{plis2014deep},
yet often only as a proof of concept.
Others are working now to try and find universal architectures for most
neuro-imaging tasks~\cite{henschel2019fastsurfer}, though it seems this is still far from a solved
application with satisfactory performances across the board.

@@

My ambition is to create a highly accurate 3d convolutional network to first classify, then segment brain scans to
help with my specific application of neuroimaging.

I may also attempt to explore feature analysis and reduction which has been shown to
help~\cite{shi2018feature} with deep learning tasks, if only to manage the challenging
dimensionality of their 3d inputs.

% ----------------------------------------------------------------------
% PROBLEM STATEMENT
% ----------------------------------------------------------------------
\section{Problem Statement}\label{sec:problem-statement}
The dataset to be used is a set of neuro-imaging scans from a cohort of 30 healthy, young women in the
Netherlands circa 2013 who were presented with various grids of images featuring
either neutral entities, (e.g.\ office chairs, windows, jackets) or foods~\cite{smeets2013allured}.
The foods were specifically types of calorically dense, tempting foods, which the individuals viewing would likely
have formed restraint with given their success in maintaining healthy body mass indices.
The experiment was meant to highlight brain structures roles in anticipated rewards from consuming delicious foods,
balanced against the hypothesized activation of restraints.
In neuro-imaging, this task is known as passive image viewing.

Given the quite important role of food and willpower in our lives, the activation patterns are likely to be powerful
and distinguishable from neutral objects which we haven't evolved to be nearly as concerned with.
This combination gives me high hopes of success in this specific application of learning to classify the viewed
objects through brain activations of the subjects.

I hope given the clarity in patterns in the brain scans, that I can develop a neural network to highlight what
features and locations in the brain scans have been helpful in informing the decisions of classification, in the
hope that such methods could be adapted to harder tasks in the future.

The dataset can be found from the excellent
OpenNeuro project here: \url{https://openneuro.org/datasets/ds000157/versions/00001}
It consists of a survey detailing the relationship of each subject with food and diet taken after the experimental
proceedings, and the core data of two modalities of brain imaging: a single high resolution T1w brain scan,
and time series of bold scans of about 370-375 scans, which encompasses the approximately 10 minute experiment
in its entirety.
The bold scans are quite sizeable given the number of them, however they are a relatively low resolution, consisting
of 64 sagital by 64 coronal by 30 axial scans.

@@

You may find this project repository here: \url{https://github.com/e-m-albright/CS682}
Reading the paper is critical to understanding the format of the dataset, however I hope I will provide enough
abstraction in the project repository for you to be freed of that challenge should you wish to have a look around.

% ----------------------------------------------------------------------
% TECHNICAL APPROACH
% ----------------------------------------------------------------------
\section{Technical Approach}\label{sec:technical-approach}
I intend to explore the efficacy of 3d convolutional networks which have been shown to have a slight performance
edge on 2d convolutions~\cite{payan2015predicting}.
Given the nature of the input images being 3d brain scans this makes intuitive sense and could allow for much more
interesting patterns of brain activity and structuring being recognized.

As for the deeper architecture, I'm hoping to imitate some of the more successful networks like VGG or ResNet,
however I wouldn't be surprised if they don't adapt well to this particular application, especially given how
interesting and important it could be to keep the input data in a three dimensional format.

It is my hope that I can also learn from other developments of object detection to see what regions in the
brain become active for a particular classification, like Faster R-CNN

\subsection{Novelty}\label{subsec:novelty}

I have not seen any application of classification or object segmentation to neuroimaging in 3d
on brain activity patterns.
I have seen some works which I hope to leverage with similar aims like that of structural identification
for tumors, in two dimensions~\cite{akkus2017deep}.
However at the very least I hope to produce a very successful network on my chosen problem first and foremost rather
than assume I am the only person to ever think of something as intuitive as bringing neural networks to bear
on neuro-imaging tasks.

I hope to find a methodology of first classifying correctly the brain scans, improve accuracy as much as possible,
and adapt the approach into tools for identifying important regions of activity in the
brain which are being utilized for classification.

I am not chiefly focused on coping with the challenges of memory but it may become too important to overlook.
Each dataset input is a $64 \times 64 \times 30$ slice of floats, for which we've got
about $375 scans \times 30 subjects$ total.
In my work on preliminary results I have already been confronted by the limitations of my current machine and have
been trying working around the issue until I am ready to attempt to resolve it.
This may be an unavoidable area to simultaneously try to innovate for the field of
neuro-imaging analysis with deep learning

% ----------------------------------------------------------------------
% INTERMEDIATE RESULTS
% ----------------------------------------------------------------------
\section{Intermediate Results}\label{sec:intermediate-results}
At the date of the milestone, I have shown that my experimental set-up is complete to begin
iterating towards the novel contributions aimed for by this project.

@@

\subsection{Sanity Checks}\label{subsec:sanity-checks}
Using the data, I have implemented a very simple, yet effective 80\% accurate SVM classifier to prove the data
should be ready for use with my chosen deep learning methodologies.

I'm now confident in the classifications of each of the scans (about 375 per subject, for 30 subjects)

\subsection{Deep Learning \& Future Works}\label{subsec:deep-learning}
I have tried feeding the data into very basic neural networks with mixed success, and am in the active process of
troubleshooting.
Once I have matched performance with the SVM using an implementation of a reasonable neural network using PyTorch, I
will begin the process of looking for refinements and innovations alluded to in previous sections, as well as
transitioning the task to object detection if I can.

% ----------------------------------------------------------------------
% TODO
% ----------------------------------------------------------------------
\section{TODO}\label{sec:todo}

NN project
- 2d hyperparameters, let's use a LR scheduler? LR spikes all over the place still, why?

- 3d CNN - typically meant for video classification using time as the third axis,
here we are not using time, though that is a dimension we have access to,
though doesn't make sense for this experiment/project to try to add in
- https://pytorch.org/docs/stable/torchvision/models.html#video-classification

- scheduler / hyperparams / diff architectures (just go with res3d, res2d+1d, mix three types
- cut batch size to avoid scaling down images (TRIED NO BUENO?)

be sure to talk about everything in paper

normalization/mean image
interpolation
wrestling with memory constraints (batch size / 3d data huge for cpu/gpu, dataset pretty large in general for ram) slow learning, tapering off learning
different styles of 1d (unfurl 3d, 2d) 2d (mean flattened, one slice) 3d (scale, orig)
LR explorations
sanity checks on brain image data, impossible for me to spot the differences

batch_size and interpolation hard to juggle too?
3d feels like "luck" plays a big part of getting training to catch/start
seems like scaling down is pretty important and I have better luck with larger batch sizes (meaning scale down is crucial) wonder if aggressive scale down is worthwhile, like over 3./4, image gets real grainy real fast)

images:
LR / accuracy measures, of diff approaches
hi res, mid res, interpolated down res brain scan examples,
2d slice example
total training time? num features comparison
average slice image
graphs from 2d training, 3d takes legit 2-3 hours for 200 epochs on 6 subjects with 5./6 & 7. / 2.

@@

image segmentation attempt?
saliency maps?
filter views?

torch save models - can eval that way
host for downloads?

@@

findings: 3d is cool, but crazy prohibitively slow and not much better than 2d without more refinements. IMHO theres
promise still in exploring it, but given the inexcusable runtime and datasize 2d is probably worth more investment
unless better hardware or serious effort is expended in getting it differentiable

@@

>>>>

\author{Anonymized for Peer Review\\
University of Massachusetts Amherst\\
140 Governors Dr, Amherst, MA 01002\\
{\tt\small @umass.edu}
}

@@

% ----------------------------------------------------------------------
% ABSTRACT
%-------------------------------------------------------------------------
\begin{abstract}
 Analyzing the brain's activity patterns starts with identifying known patterns of well established structures,
 typically observed by human oversight in two dimensions, which may miss subtle, barely visible but useful indicators,
 and limit the ability to stitch together patterns across a third dimension.

 Deep learning's ability to decipher and learn from such nearly impossible patterns in massive datasets,
 imperceptible to preliminary observation makes it an excellent candidate to explore and teach us about
 overlooked patterns of activity.

 Here I showcase convolutional network's ability to correctly classify uncleaned activity data from a passive
 image viewing experiment aimed at triggering activity in particular regions of the brain controlling reward and
 restraint, using the data in both 2d and 3d to explore the merits of adapting video-processing techniques meant
 to apply across time to a third spatial dimension and discuss adapting this to unsupervised segmentation to see
 what the network can teach us from its findings.

\end{abstract}

@@

Classification of activity and structures should come quite naturally with deep learning practices.
Previous works include the development of Restricted Boltzmann Machines, and Deep Belief Networks~\cite{plis2014deep},
yet often only as a proof of concept, and limited to two dimensions.
Others are working now to try and find universal architectures for most
neuro-imaging tasks~\cite{henschel2019fastsurfer}, though it seems this is still far from a solved
application with satisfactory performances across the board.

@@

Here we'll discuss creating highly accurate classifiers using both 2d and 3d convolutional networks
to learn in the typical manner low level filters, here specifically created for brain scans of spatial activity, using
no pretraining on unrelated datasets and no neuro-imaging feature analysis or reduction~\cite{shi2018feature},
letting the model learn for itself what is important so that it can teach the observer instead.

Following this we'll discuss what this may mean for future works in the highly valuable ability to adapt the classifiers
to the new, related task of unsupervised segmentation in 3 dimensions~\cite{shu2016unsupervised} to mirror what has been
a fruitful field of 2d identification of tumors~\cite{akkus2017deep} for this specific application of telling brain
activity patterns.

% ----------------------------------------------------------------------
% BACKGROUND / RELATED
% ----------------------------------------------------------------------
\section{Background}\label{sec:background}

The dataset to be used is a set of neuro-imaging scans from a cohort of 30 healthy, young women in the
Netherlands circa 2013 who were presented with various grids of images featuring
either neutral entities, (e.g.\ office chairs, windows, jackets) or foods~\cite{smeets2013allured}.
The foods were specifically types of calorically dense, tempting foods, which the individuals viewing would likely
have formed restraint with given their success in maintaining healthy body mass indices.
The experiment was meant to highlight brain structures roles in anticipated rewards from consuming delicious foods,
balanced against the hypothesized activation of restraints.

In neuro-imaging, this task is known as passive image viewing.

Given the quite important role of food and willpower in our lives, the activation patterns are likely to be powerful
and distinguishable from neutral objects which we haven't evolved to be nearly as concerned with.
This combination makes this an excellent selection for a proof of concept application of deep learning as the chances
of success on a less subtle activation pattern, observable to trained researchers, are quite high.
This specific application then can serve as a jumping off point for future works where the activity patterns are less
distinguishable, and may be an interesting addition to the works in brain-pattern controlled prosthetics.
This dataset is also ideal as it comes with a fairly usable ground-truth labeling mechanism, as I am not currently
investigating the capacity for unsupervised differentiation.

The dataset can be found from the excellent
OpenNeuro project here: \url{https://openneuro.org/datasets/ds000157/versions/00001}

The dataset follows the BIDS (Brain Imaging Data Structure) layout adopted by many neuroscience researchers.
It consists of a survey detailing the relationship of each of the 30 subjects with food and diet taken after the experimental
proceedings, and the core data of two modalities of brain imaging: a single high resolution T1w Modality fMRI scan seen in figure~\ref{fig:ortho-t1w},
and time series of 370-375 BOLD Modality fMRI scans seen in figure~\ref{fig:ortho-bold}, encompassing approximately 10 minutes
capturing brain activity of the subjects during their exposure to the experimental triggers.

The bold scans are quite sizeable as far as datasets go, given the number of each per subject,
however they are a relatively low resolution, consisting
of 64 sagital by 64 coronal by 30 axial scans, with only a single image channel.

The activation is measured in a positive magnitude visible in figure~\ref{fig:ex_stat_map}, through a helper method in a neuro-imaging specialized
image handling library ~\cite{brett_matthew_2019_3544468}.

I had to manually infer pairings of the scans with the experimental phase for classification training given the events log associated
with each subject and the advertised scan times from the paper.
While the data does appear healthy, there may be some discrepancies of interpretation which would warrant future investigation if the
performance on this particular experiment was the primary focus.

@@

You may find this project repository here: (personal github link removed for anonymity)
Reading the paper is critical to understanding the format of the dataset, however I hope I will provide enough
abstraction in the project repository for you to be freed of that challenge should you wish to have a look around.

\begin{figure}
  \includegraphics[width=\linewidth]{images/orthoview_bold.png}
  \caption{Bold Modality Low-Res 3d Experimental Scans of Subject Brain}
  \label{fig:ortho-bold}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{images/example_stat_map.png}
  \caption{Activity Measures from Bold Scans}
  \label{fig:ex_stat_map}
\end{figure}

% ----------------------------------------------------------------------
% APPROACH
% ----------------------------------------------------------------------
\section{Approach}\label{sec:approach}

Beginning with SKLearn~\cite{scikit-learn} data readiness was proven with moderately successful classification using
a simple SVM and then the SKLearn Multi-Layer-Perceptron neural network, getting 70-80\% accuracy, sufficing the
initial sanity check that deep learning would be ready to work on.

% TODO these are relevant citations, right?
Leveraging PyTorch~\cite{paszke2017automatic} first a framework for learning was likewise vetted using a simple fully
connected shallow network, graduating to the state of the art convolutional networks, chiefly focussed
on the ResNet~\cite{he2016deep} architecture styles, which have been developed for video classification~\cite{tran2018closer}
utilizing 3 dimensional convolutions~\cite{hara3dcnns} much in the way 2 dimensional convolutions operate, which have
been shown to hold a slight edge on 2 dimensions~\cite{payan2015predicting} in other applications.

\subsection{Data Preparation}\label{subsec:data-preparation}

First as the data was not prepared for classification, critical cleaning include subtracting the mean image to better
highlight meaningful variance, and standardizing the data between 0 and 1.

Given the nature of the input images being 3d brain scans it's intuitive that examining the brain in 3 dimensions
allows for much more interesting patterns of brain activity and structuring being recognized.

This comes with a substantial cost however, as the extra dimension explodes the memory demands of the input data, and
the number of trainable parameters for the models.
Many of the recommendations following will vary based on the particular hardware, but with a 4GB GPU I had to be quite
careful about batch-sizes to not over-allocate memory.
To combat this, first, a relatively conservative amount of blank space is cropped from the images, halving the total footprint.
Following this, in order to use a batch-size of 64, I interpolate images to approximately $\frac{5}{6}$ their original dimension

In addition, I typically do not use the full cohort of 30 subjects, as I even encounter some RAM restrictions with the
full raw dataset consisting of
$64 sagital \times 64 coronal \times 30 axial \times 32bit floats \times 375 scans \times 30 subjects$ or roughly 5.53 gigabytes

In working with 2 dimensions, omitting the axial dimension for a flat cross section, you'll see the dataset is
$\frac{1}{30}$ the size and hence, much more manageable.

\subsection{Examining the Data}\label{subsec:examining-data}

Two fully processed, correctly classified brain scans are presented in the figure~\ref{fig:comparison}
which pose a daunting challenge to human labelers.

\begin{figure}
  \includegraphics[width=\linewidth]{images/nf_f_hard_comparison.png}
  \caption{Nonfood-Stimuli \& Food-Stimuli Respectively from the same Subject}
  \label{fig:comparison}
\end{figure}

\subsection{Model Selection}\label{subsec:model-selection}

For the primary focus of comparing and improving classification accuracy using 2d and 3d architectures, ResNet models
of the standard 18 layers to 100+ layers are examined, as the ResNet family is both highly competitive, and demonstrated
in both 2 and 3 dimensions which will help comparisons in exploring the benefits of 3d brain scan usage.
Video processing typically limits the layers to a more shallow level than that of image processing, as unfortunately as
with the data, the model parameters explode out of control pretty fast.
The following parameter counts are from standard ResNet architectures adapted to a single input channel.

\begin{center}
 \begin{tabular}{||c c c||}
 \hline
 - & ResNet-18 2d & ResNet-18 3d \\ [0.5ex]
 \hline\hline
 Counts & 11,171,266 & 33,148,482 \\ [1ex]
 \hline
 Ratio & 1.0 & 2.97 \\ [1ex]
 \hline
\end{tabular}
\end{center}

\subsection{Comparisons}\label{subsec:comparisons}

Model performance is typically evaluated using accuracy, precision, recall, and AUC metrics, as well as total time to
train.

% ----------------------------------------------------------------------
% RESULTS
% ----------------------------------------------------------------------
\section{Results}\label{sec:results}
At the date of the milestone, I have shown that my experimental set-up is complete to begin
iterating towards the novel contributions aimed for by this project.

@@

\subsection{Model Architecture}\label{subsec:model-arch}

the project explored minimally the hyperparameters of the chosen model architectures
chiefly was finding the right learning rate for the given normalized data to the model
typically 1e-2 to 1e-4 worked, with 1e-3 being good, and decay being very very helpful
model quite sensitive to adjustments

- scheduler / hyperparams / diff architectures (just go with res3d, res2d+1d, mix three types

\subsection{Scope of Number of Subjects}\label{subsec:num-subjects}

@@

 \begin{figure}
  \includegraphics[width=\linewidth]{images/2d_cold.png}
  \caption{Cold-Start 2D Training Progress}
  \label{fig:2d_cold}
\end{figure}

\subsection{Warm Starts}\label{subsec:warm-starts}
saving model parameter dict, loading and training again
wowee good stuff, hope to use a LR scheduler in the future to try to control this better?
surprisingly bad results on warm starting expansion to new subjects - old same subjects takes a little while to stabilize
but it gets back to 83% without much fuss, probably LR is a little high to start with and the decay helps a lot.
bounces down for a while, bounces up for a while. breaking 90% which is very cool, might mean warm start has best potential

 \begin{figure}
  \includegraphics[width=\linewidth]{images/2d_warm.png}
  \caption{Warm-Start 2D Training Progress}
  \label{fig:2d_warm}
\end{figure}

\subsection{3 Dimensions}\label{subsec:3-dimensions}
- 3d CNN - typically meant for video classification using time as the third axis,
here we are not using time, though that is a dimension we have access to,
though doesn't make sense for this experiment/project to try to add in
- https://pytorch.org/docs/stable/torchvision/models.html#video-classification

okay it works, and has potential, certainly given the data is richer to be the best, but wow it's frustratingly slow
findings: 3d is cool, but crazy prohibitively slow and not much better than 2d without more refinements. IMHO theres
promise still in exploring it, but given the inexcusable runtime and datasize 2d is probably worth more investment
unless better hardware or serious effort is expended in getting it differentiable

 \begin{figure}
  \includegraphics[width=\linewidth]{images/3d_cold.png}
  \caption{Cold-Start 3D Training Progress}
  \label{fig:3d_cold}
\end{figure}
wrestling with memory constraints (batch size / 3d data huge for cpu/gpu, dataset pretty large in general for ram) slow learning, tapering off learning
different styles of 1d (unfurl 3d, 2d) 2d (mean flattened, one slice) 3d (scale, orig)
LR explorations

@@

"none of my alterations to the model architecture worked that well, most did nothing, a few broke things, could still poke around more though"
graphs from 2d training, 3d takes legit 2-3 hours for 200 epochs on 6 subjects with 5./6 & 7. / 2.
batch_size and interpolation hard to juggle too?
3d feels like "luck" plays a big part of getting training to catch/start
seems like scaling down is pretty important and I have better luck with larger batch sizes (meaning scale down is crucial) wonder if aggressive scale down is worthwhile, like over 3./4, image gets real grainy real fast)

cut batch size to avoid scaling down images (TRIED NO BUENO?)

\subsection{Performance}\label{subsec:performance}

 % P-R-F1-S
\begin{center}
 \begin{tabular}{||c c c||}
 \hline
 - & Nonfood & Food \\ [0.5ex]
 \hline\hline
Precision & 0.949074 & 0.958904 \\
 \hline
Recall & 0.957944 & 0.950226 \\
 \hline
F1 & 0.953488 & 0.954545 \\
 \hline
Support & 214 & 221 \\ [1ex]
 \hline
\end{tabular}

 % ROC-AUC / ACC
\end{center}
\begin{center}
 \begin{tabular}{||c c||}
 \hline
 ROC-AUC & Accuracy \\ [0.5ex]
 \hline\hline
  0.9847 & 0.954 \\ [1ex]
 \hline
\end{tabular}
\end{center}

Can get a little higher but these are already encouraging numbers (96 seen) often the model starts leaning a little
one way or the other and this was one of the most balanced runs I've seen

I'm a little unclear about the reliability of my data labeling, so I'm happy with this result, I'd be stoked
to reproduce on 3d, and trusting the model this much, open the door to segmentation

% ----------------------------------------------------------------------
% DISCUSSION
% ----------------------------------------------------------------------
\section{Discussion}\label{sec:discussion}

\subsection{Learning}\label{subsec:learning}
very cool! I tried spot checking different scans to see if I could tell the difference visually and I really really
cannot. I think it's encouraging that the model is able to notice impossible to visualize differences and run with it
this well. It's one thing to differentiate a st. bernard from a siamese cat, but two brain activations I'm not good at

\subsection{Future Works}\label{subsec:future-works}
Unsupervised segmentation, especially in 3d would be cool
Future works
- could strive to refine 3d more, with more time and compute
- unsupervised segmentation would be cool
- find ways to not have to smoosh images for memory?
- more visualizations? lack of color kinda sucks, saliency is weird, filters prob too

began work on visualization, it didn't lead anywhere useful, 1 channel might be a problem?
- check out filter viz
- see about unsupervised
- try 3d alternative archs? (dont really need to)

might want to improve 3d before going on to segmentation,

 \begin{figure}
  \includegraphics[width=\linewidth]{images/saliency.png}
  \caption{caption}
  \label{fig:saliency}
\end{figure}

@@

% ----------------------------------------------------------------------
% SUPPLEMENTARY
% ----------------------------------------------------------------------
\section{Supplementary Material}\label{sec:supplementary}

Extra material for reference

 \begin{figure}
  \includegraphics[width=\linewidth]{images/orthoview_t1w.png}
  \caption{caption}
  \label{fig:ortho-t1w}
\end{figure}

 \begin{figure}
  \includegraphics[width=\linewidth]{images/var_on_cortex_2.png}
  \caption{caption}
  \label{fig:cortex-var}
\end{figure}
